Product plots

### 2011-06-22 prodplots-infovis.tex
extended discussion on continuous variables, started related work section, fixed some typos

\title{Product plots}
\author{Hadley Wickham, and Heike Hofmann}
\authorfooter{
  \item Hadley Wickham is an Assistant Professor of Statistics at Rice University, Email: hadley@rice.edu.
  \item Heike Hofmann is an Associate Professor of Statistics at Iowa State University.
}
\shortauthortitle{Wickham \MakeLowercase{\textit{et al.}}: Product plots}

@@

Section~\ref{sec:combination} provides the mathematical framework that allows us to combine the 1d and 2d atoms to display data of any dimensionality. This framework is based on the fundamental statistical idea that high any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions. Section~\ref{sec:existing} shows how many existing named graphics are special cases of this general pattern.

@@

@@

\noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning! Well thought out interaction can make it easy to navigate this space. We have a prototype interaction model developed as part of the cranvas suite of interactive graphics in R, {\tt http://github.com/ggobi/cranvas}, but much work and user testing remains to be done.

We also are hopeful that developing the connection between product plots and log-linear (aka Poisson) models will also help better understand this space of plots. Some special cases have already been worked out. For example, \citep{hofmann:2003} shows how looking for straight lines in a mosaic plot corresponds to a formal test of independence between two variables. Generally, what visual comparisons are equivalent to what formal statistical tests? And conversely, how can inspecting the significant coefficients in a model help us choose a set of useful plots for a given dataset? Generally, given a question, can we suggest appropriate plots? Or given a plot, can we suggest questions that it might answer?

>>>>

\title{Product plots}
\author{Hadley Wickham, and Heike Hofmann}
\authorfooter{
  \item Hadley Wickham is Assistant Professor of Statistics at Rice University, Email: hadley@rice.edu.
  \item Heike Hofmann is Associate Professor of Statistics at Iowa State University.
}
\shortauthortitle{Wickham \MakeLowercase{\textit{et al.}}: Product plots}

@@

Section~\ref{sec:combination} provides the mathematical framework that allows us to combine the 1d and 2d atoms to display data of any dimensionality. This framework is based on the fundamental statistical idea that any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions. Section~\ref{sec:existing} shows how many existing named graphics are special cases of this general pattern.

@@

A more theoretical approach in dealing with continuous variables  lies in increasing the number of bins infinitesimally -- this leads from a probability mass function to a density function, making the 1d primitives of hbar and vbar to density plots, and changing the spine-plot representation to a conditional density plot (citation needed - probably Martin's and Heike's paper). While the inclusion of one continuous variable is completely feasible that way, two variables need the use of an additional aesthetic, such a s colour,  to visualize a 2d joint density. Three or more continuous variables break the scope of the product plots framework.

@@

\noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning! Well thought out interaction can make it easier to navigate this space. We have a prototype interaction model developed as part of the cranvas suite of interactive graphics in R, {\tt http://github.com/ggobi/cranvas}, but much work and user testing remains to be done.

We also are hopeful that developing the connection between product plots and log-linear (aka Poisson) models will also help better understand this space of plots. Some special cases have already been worked out. For example, \citep{hofmann:2003} shows how looking for straight lines in a mosaic plot corresponds to a formal test of independence between two variables. Generally, what visual comparisons are equivalent to what formal statistical tests? And conversely, how can inspecting the significant coefficients in a model help us choose a set of useful plots for a given dataset? Can we, given a question,  suggest appropriate plots? Or given a plot, can we suggest questions that it might answer?
